Alternating Automata Modulo First Order Theories

### 2019-02-07 quantifiers/appendix.tex
ok

>>>>

\subsection{Proof of Lemma \ref{lemma:refinement}}

The new set of formulae $U'(\alpha_0), \ldots, U'(\alpha_n)$ complies
with Definition \ref{def:unfolding}, because: \begin{compactitem}
\item $U'(\alpha_0) \equiv \iota$, since, by point
  \ref{it2:generalized-lyndon-interpolant} of Definition
  \ref{def:generalized-lyndon-interpolant}, we have $\stamp{\iota}{0}
  \models I_0$, thus $\iota \models J_0$ and $U'(\alpha_0) =
  U(\alpha_0) \wedge J_0 \equiv \iota \wedge J_0 \equiv \iota$, and
    %
\item by Definition \ref{def:generalized-lyndon-interpolant}
  (\ref{it3:generalized-lyndon-interpolant}) we have, for all $k \in
  [n-1]$:
  \[I_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y})
  \rightarrow \stamp{\psi}{k+1} \models I_{k+1}\] We write
  $\overstamp{I}{j}_k$ for the formula in which each predicate
  symbol $\stamp{q}{k}$ is replaced by $\stamp{q}{j}$. Then the
  following entailment holds:
  \[\overstamp{I}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
  \rightarrow \stamp{\psi}{1} \models \overstamp{I}{1}_{k+1}\]
  Because $J_k$ is obtained by removing the time stamps from the predicate symbols and  
  existentially quantifying all the free variables of $I_k$, we also
  obtain, applying Fact \ref{fact:ex-entail} below:
  \[\stamp{J}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
  \rightarrow \stamp{\psi}{1} \models \stamp{J}{1}_{k+1}\] Since $U$
  satisfies the labeling condition of Definition \ref{def:unfolding}
  and $U'(\alpha_k) = U(\alpha_k) \wedge J_k$, we obtain, as required:
  \[\stamp{U'(\alpha_k)}{0} \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
  \rightarrow \stamp{\psi}{1} \models
  \stamp{U'(\alpha_{k+1})}{1}\enspace.\]
\end{compactitem}

\begin{fact}\label{fact:ex-entail}
  Given formulae $\phi(\vec{x}, \vec{y})$ and $\psi(\vec{x})$ such
  that $\phi(\vec{x}, \vec{y}) \models \psi(\vec{x})$, we also have
  $\exists \vec{x} ~.~ \phi(\vec{x},\vec{y}) \models \exists \vec{x}
  ~.~ \psi(\vec{x})$. 
\end{fact}
\paragraph{\em Proof.} For each choice of a valuation for the existentially
quantified variables on the left-hand side, we chose the same
valuation for the variables on the right-hand side. \qed \qed

### 2019-02-07 quantifiers/body.tex
ok

Consider, for instance the problem of proving that $\lang{A} \subseteq
\lang{B}$, where $A$ and $B$ are the timed automata in
Fig. \ref{fig:ta}. The automaton $A$ ($B$) has a single clock $x$
($y$) which is reset on the transition from $s_1$ to $s_2$ ($q_1$ to
$q_2$). Intuititvely, $A$ recognizes the sequences in which all events
are $a$ and the second event, if it occurs at all, it occurs after at
least $2$ time units. On the other hand, $B$ recognizes the sequences
in which there exists an event $a$ such that no event occurs after it
in exactly $1$ time unit. Formally, $\lang{A} = \set{a, t_1, \ldots a,
  t_n \mid t_1, \ldots, t_n \in \real,~ n > 1 \Rightarrow t_2-t_1 \geq
  2}$ and $\lang{B} = \set{\sigma_1, t_1 \ldots \sigma_{i-1}, t_{i-1},
  a, t_i, \sigma_{i+1}, t_{i+1}, \ldots \sigma_n, t_n \mid \sigma_1,
  \ldots, \sigma_n \in \set{a,b},~ n \geq 1,~ t_1, \ldots, t_n \in
  \real,~ 1 \leq i < n \Rightarrow t_{i+1}-t_i \neq 1}$. It is
proved\footnote{See \cite[Theorem 1]{AlurMadhusudan04} for a proof.}
that $\clang{B}$ cannot be recognized by a timed automaton, thus the
inclusion cannot be proved in the framework of timed regular
languages, as $\lang{A} \cap \clang{B} = \emptyset$.

Our approach is to encode timed automata as first order alternating
automata and use closure of the latter class under intersection and
complement to compute an alternating automaton, whose emptiness proves
that $\lang{A} \subseteq \lang{B}$. Consider the encoding of $A$ in
Fig. \ref{fig:ta} (right) inspired by a translation of timed automata
into Datalog programs \cite{Fribourg98}.  Each control state of $A$
and $B$ is represented by a binary predicate symbol with the same
name. The first argument of $s_1(x,z)$ tracks the last moment in time
when the $\mathsf{x}$ clock of $A$ was reset and $z$ is the current
time value. The time stamp of the input event $a$ is the input
variable $t$, which must be bigger or equal to $z$, since the lapse of
time between two consequtive events is always positive. The current
value of the $\mathsf{x}$ clock is $z-x$, thus the guard $\mathsf{x}
\geq 2$ on the $s_2 \arrow{}{} s_2$ transition of $A$ from becomes
$z-x \geq 2$ in the second rule, and the reset of $x$ on the $s_1
\arrow{}{} s_2$ transition is encoded by passing $z$ as the first
parameter of $s_2$, in the first rule. The initial configuration of
$A$ [$B$] is given by the atomic propositon $s_1(0,0)$ [$q_1(0,0)$]
and $s_2$ [$q_2$] marks the final configurations.

@@

The first order alternating automaton recognizing the (encoding of
the) $\clang{B}$ language is simply obtained by negating the theory
atoms $t \geq u$ and $u - y \neq 1$ and interchanging conjunctions
with disjunctions in the rules of the alternating automaton encoding
$B$, denoted as $\overline{B}$ in Fig. \ref{fig:ta} (bottom right),
with final configurations marked by $q_1$, instead of $q_2$. Finally,
we check the emptiness of the alternating automaton consisting of the
rules of $A$ and $\overline{B}$, started in the initial configuration
$s_1(0,0) \wedge q_1(0,0)$. Our lazy annotation semi-algorithm proves
emptiness of this automaton in $\sim\!0.3$ seconds, on an average
laptop machine. The missing proofs of the technical results are given
in Appendinx \ref{app:proofs}.

@@

\section{Preliminaries}

@@

\subsection{Closure Properties}
\label{sec:closure}

>>>>

Consider, for instance the problem of proving that $\lang{A} \subseteq
\lang{B}$, where $A$ and $B$ are the timed automata in
Fig. \ref{fig:ta}. The automaton $A$ ($B$) has a single clock $x$
($y$) which is reset on the transition from $s_1$ to $s_2$ ($q_1$ to
$q_2$). Intuititvely, $A$ recognizes the sequences in which all events
are $a$ and the second event, if it occurs at all, it occurs after at
least $2$ time units. On the other hand, $B$ recognizes the sequences
in which there exists an event $a$ such that no event occurs after it
in exactly $1$ time unit. Formally, $\lang{A} = \set{a, t_1, \ldots a,
  t_n \mid t_1, \ldots, t_n \in \real,~ n > 1 \Rightarrow t_2-t_1 \geq
  2}$ and $\lang{B} = \set{\sigma_1, t_1 \ldots \sigma_{i-1}, t_{i-1},
  a, t_i, \sigma_{i+1}, t_{i+1}, \ldots \sigma_n, t_n \mid \sigma_1,
  \ldots, \sigma_n \in \set{a,b},~ n \geq 1,~ t_1, \ldots, t_n \in
  \real,~ 1 \leq i < n \Rightarrow t_{i+1}-t_i \neq 1}$. It is
proved\footnote{See \cite[Theorem 1]{AlurMadhusudan04} for a proof.}
that $\clang{B}$ cannot be recognized by a timed automaton, thus the
inclusion cannot be proved in the framework of timed automata
\cite{AlurDill94,AlurMadhusudan04}, by showing that $\lang{A} \cap
\clang{B} = \emptyset$.

Our approach is to encode timed automata as first order alternating
automata and use closure of the latter class under intersection and
complement to compute an alternating automaton, whose emptiness proves
that $\lang{A} \subseteq \lang{B}$. Consider the encoding of $A$ in
Fig. \ref{fig:ta} (right) inspired by a translation of timed automata
into Datalog programs \cite{Fribourg98}.  Each control state of $A$
and $B$ is represented by a binary predicate symbol with the same
name. The first argument of $s_1(x,z)$ tracks the (last) time value
time when the $\mathsf{x}$ clock of $A$ was reset (if it is reset
more than once) and $z$ is the time value of the last event. The time
stamp of the current event $a$ is the input variable $t$, which must
be bigger or equal to $z$, since the lapse of time between two
consequtive events is always positive. The current value of the
$\mathsf{x}$ clock is thus $z-x$ and the guard $\mathsf{x} \geq 2$ on
the $s_2 \arrow{}{} s_2$ transition of $A$ from becomes $z-x \geq 2$
in the second rule. The reset of $x$ on the $s_1 \arrow{}{} s_2$
transition is encoded by passing $z$ as the first parameter of $s_2$,
in the first rule. The initial configuration of $A$ [$B$] is
$s_1(0,0)$ [$q_1(0,0)$] and $s_2$ [$q_2$] marks the final
configurations.

@@

The first order alternating automaton recognizing the (encoding of
the) $\clang{B}$ language is simply obtained by negating the theory
atoms $t \geq u$ and $u - y \neq 1$ and interchanging conjunctions
with disjunctions in the rules of the alternating automaton encoding
$B$, denoted as $\overline{B}$ in Fig. \ref{fig:ta} (bottom right),
with final configurations marked by $q_1$, instead of $q_2$. Finally,
we check the emptiness of the alternating automaton consisting of the
rules of $A$ and $\overline{B}$, started in the initial configuration
$s_1(0,0) \wedge q_1(0,0)$. Our lazy annotation semi-algorithm proves
emptiness of this automaton in $\sim\!0.3$ seconds, on an average
laptop machine.

@@

For space reasons, the proofs of the technical results are given in
Appendinx \ref{app:proofs}.

\vspace*{-.5\baselineskip}
\subsection{Preliminaries}

@@

\subsection{Boolean Closure of First Order Alternating Automata}
\label{sec:closure}
